The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-08-22 paper.qmd
read over + edited visualising uncertainty as a variable

## The choropleth map
@fig-choropleth depicts a choropleth map of the counties of Iowa. Each of these counties are colored according to an estimate of average daily temperature that was simulated so that the values followed a clear spatial trend (hot in the middle of the map, and cold on the outside). The variance of these estimates were simulated such that a hypothesis test would indicate the existence of a spatial trend in the low variance map, while the trend in the high variance map should be indistinguishable from noise. Is this distinction in validity of the spatial trend clear in in the map? Is the validity of the trend communicated through the visualisation?

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-choropleth
#| fig-cap: "Two choropleth maps that depict the counties of Iowa where each country coloured acording to a simulated average temperature. Both maps depict a spatial trend, where counties closer to the center of the map are hotter than counties on the edge of the map. This trend is only technically there in the low variance condition, i.e. if we performed a hypothesis test we would conclude the trend exists. The high variance condition displays a spatial trend that could simply be the result of noise, which means the plot is displaying a false conclusion."
#| fig-subcap: 
#|   - "Low Variance Data"
#|   - "High Variance Data"
#|   - "Choropleth Palette"
#| layout-ncol: 3
#| layout-valign: "bottom"
#| cap-location: "bottom"

@@

## Signal-supression
The two choropleth maps appearing to be identical in @fig-choropleth highlights the need for uncertainty visualisations. Uncertainty visualisation is is required for transparency and this sentiment has been repeated.  Some authors suggest uncertainty is important to include as it communicates the legitimacy (or illegitimacy) of the conclusion drawn from visual inference [@Correll2014; @Kale2018; @Griethe2006]. Some authors have said that uncertainty should be included to degree of confidence or trust in the data [@Boukhelifa2012; @Zhao2023]. Some authors directly connect uncertainty visualisation to hypothesis testing as it ensures the "validity" of a statement [@Hullman2020a; @Griethe2006], but allows for a proportional level of trust that is more detailed than the binary results of a hypothesis test [@Correll2014; @Correll2018]. Some authors even go so far as to claim that failing to include uncertainty is akin to fraud or lying [@Hullman2020a; @Manski2020].

This consensus leads us to understand that uncertainty visualisation is motivated by the need for a sort of "visual hypothesis test". A successful uncertainty visualisation would act as a "statistical hedge" for any inference me make using the graphic. Additionally, since the purpose of a visualisation is to give a quick "gist" of the information [@Spiegelhalter2017], this hedging needs to be communicated visually without any need for computation from the viewer. Additionally, @Ndlovu2023 found that participants applied the same methods they used for simple choropleth maps to complicated uncertainty maps even if that take away was invalid. Therefore, this hedging effect needs to be communicated simply through the visualisation. If we refer to the conclusion we draw from a graphic to be its "signal" and the variance that makes this signal harder to identify as the "noise", we can summaries this information into three key requirements. A good uncertainty visualisation needs to:

  1) Reinforce justified signals to encourage confidence in results
  2) Hide signals that are just noise to prevent unjustified conclusions
  3) Perform tasks 1) and 2) in a way that is proportional to the level of confidence in those conclusions.

As @fig-choropleth showed, visualisations that are unconcerned with uncertainty have no issue showing justified signals, but struggle with the display of unjustified signals. Therefore, we coin this approach to uncertainty visualisation as "signal-suppression" since it primarily differentiates itself from from the "noiseless" visualisation approach through criteria (2). That is, the main difference between an uncertainty visualisation and a "normal" visualisation is that an uncertainty visualisation should prevent us from drawing unjustified conclusions.

## Uncertainty as a signal
Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss to ignore these alternative motivations. Some authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017]. Other authors do not describe uncertainty as important for decision making, but rather explicitly state it as variable of importance in of itself [@Blenkinsop2000]. While uncertainty can provide useful information in decision making, it is important to recognize the "uncertainty" in these cases is not acting as "uncertainty" at all. It is acting as signal.

This is obvious for the cases where we are explicitly interested in the variance or error, as we are literally trying to draw conclusions about an "uncertainty" statistic. The same is true for "decision making" experiments, but it is less overt. This is easiest to understand with an example. Imagine you are trying to decide if you want to bring an umbrella with you to work. An umbrella is annoying to bring with you, so you only want to pack it if the chance of rain is greater than 10%. Unfortunately, your weather prediction app only provides you with the predicted daily rainfall. Therefore, your decision will be improved with the inclusion of uncertainty, *not* because uncertainty is important for your decision, but because it gives you the tools required to calculate the *actual* statistic you are basing your decision on. In this sense, uncertainty is no more "special" to decision making than weight is in a BMI calculation.

Uncertainty visualisation's made for these purposes should simply display the uncertainty statistic we are interested in, such as the variance, or probability of an event. This is precisely what we observe. @fig-exceed depicts an exceedance probability map that has was designed as an alternative to the choropleth map to improve decision making under uncertainty [@Kuhnert2018; @Lucchesi2021]. A keen viewer may notice that the "exceedance probability map" is actually just a choropleth map, only the statistic being displayed has changed. We do not believe this graphic be considered an uncertainty visualisation.

@@

There seem to be two different definitions of "uncertainty visualisation" floating around in the literature. The first considers *any* visualisation of error, variance, or probability to be an uncertainty visualisation. The second believes an "uncertainty visualisation" is the output of a function that takes a normal visualisation as an input, and transforms it to include the uncertainty information. The former group believe the purpose of uncertainty visualisation to provide signal about a distribution, while the later believe it should act as noise to obfuscate a signal. The lack of explicit distinction between these two motivations leaves the literature muddled and reviewers struggle to understand if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014]. This disagreement creates constant contradictions in what the literature considers to be an "uncertainty visualisation". For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, and @Wickham2011 suggests their product plot framework, which includes histograms and bar charts, should be extended to include uncertainty however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. If you view uncertainty as a function applied to an existing graphic, then you would believe a pie chart and bar chart are not uncertainty visualisations, as they are yet to have the "uncertainty visualisation function" applied to them. If you view uncertainty as any graphic that depicts an "uncertainty statistic" then there are no limitations on which graphics can or cannot be uncertainty visualisations.

@@

It is clear that the pixel-map is not only "suppressing" the false information, but it is doing so by simulating *more* information. The efficacy of this method means that visualisations of simulated samples pop up repeatedly in the literature, with examples including samples that are animated over time [@Hullman2015; @Blenkinsop2000], pixel-maps, and spaghetti time series plots. Not only does this method help readers understand the plot level "gist", it is also unlikely to damage the viewers ability to extract individual estimates. Extracting global statistics, such as the mean or variance, from a sample can be done with relative each, especially when those values are mapped to color [@Franconeri2021]. Therefore, the pixel map performs signal suppression, without sacrificing the viewers ability to extract general statistics, unless those statistics *should* be harder to extract due to the uncertainty in the value. So is this the best uncertainty visualisation? And if so, why?

>>>>

## The choropleth map
@fig-choropleth depicts a choropleth map of the counties of Iowa. Each of these counties are colored according to an estimate of average daily temperature that was generated so that the values followed a clear spatial trend (hot in the middle of the map, and cold on the outside). The variance of these estimates were simulated such that the trend accounts for most of the variance in the plot in the low variance case (so we should expect the trend to be visible) while the high variance case has significantly more noise, so much so that it overwhelms the spatial trend. Is this aspect of the data and the spatial data it communicates clear in in the map? Is the strength of the trend communicated through the visualisation?

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-choropleth
#| fig-cap: "Two choropleth maps that depict the counties of Iowa where each country coloured acording to a simulated average temperature. Both maps depict a spatial trend, where counties closer to the center of the map are hotter than counties on the edge of the map. In the low variance condition, the trend accounts for most of the variation in the data, in the high variance case, the variance on the temperature estimate accounts for most of the variance. The high variance condition displays a spatial trend that could simply be spurious, which means the plot is displaying a false conclusion."
#| fig-subcap: 
#|   - "Low Variance Data"
#|   - "High Variance Data"
#|   - "Choropleth Palette"
#| layout-ncol: 3
#| layout-valign: "bottom"
#| cap-location: "bottom"

@@

## Signal-supression
Uncertainty visualisation is required for transparency. The two choropleth maps that appear to be identical in @fig-choropleth highlights the issues with simply electing to ignore uncertainty. This sentiment appears frequently in the uncertainty visualisation literature. Some authors suggest uncertainty is important to include as it communicates the legitimacy (or illegitimacy) of the conclusion drawn from visual inference [@Correll2014; @Kale2018; @Griethe2006]. Some authors have said that uncertainty should be included to degree of confidence or trust in the data [@Boukhelifa2012; @Zhao2023]. Some authors directly connect uncertainty visualisation to hypothesis testing as it ensures the "validity" of a statement [@Hullman2020a; @Griethe2006], but allows for a proportional level of trust that is more detailed than the binary results of a hypothesis test [@Correll2014; @Correll2018]. Some authors even go so far as to claim that failing to include uncertainty is akin to fraud or lying [@Hullman2020a; @Manski2020].

This consensus leads us to understand that uncertainty visualisation is motivated by the need for a sort of "visual hypothesis test". A successful uncertainty visualisation would act as a "statistical hedge" for any inference me make using the graphic. Additionally, since the purpose of a visualisation is to give a quick "gist" of the information [@Spiegelhalter2017], this hedging needs to be communicated visually without the need for complicated calculations. If we refer to the conclusion we draw from a graphic to be its "signal" and the variance that makes this signal harder to identify as the "noise", we can summaries the above information into three key requirements. A good uncertainty visualisation needs to:

  1) Reinforce justified signals to encourage confidence in results
  2) Hide signals that are overwhelmed by noise to prevent unjustified conclusions
  3) Perform tasks 1) and 2) in a way that is proportional to the level of confidence in those conclusions.

As @fig-choropleth showed, visualisations that are unconcerned with uncertainty have no issue showing justified signals, but struggle with the display of unjustified signals. Therefore, we call this approach to uncertainty visualisation as "signal-suppression" since it primarily differentiates itself from from the "noiseless" visualisation approach through criteria (2). That is, the main difference between an uncertainty visualisation and a "normal" visualisation is that an uncertainty visualisation should prevent us from drawing unjustified conclusions.

## Uncertainty as a signal
Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss not to mention these alternative approaches. Some authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017]. Other authors do not describe uncertainty as important for decision making, but rather explicitly state it as variable of importance in of itself [@Blenkinsop2000]. While uncertainty can provide useful information in decision making, it is important to recognize that the "uncertainty" in these cases is not acting as "uncertainty" at all. It is acting as signal.

This is obvious for the cases where we are explicitly interested in the variance or error, as we are literally trying to draw conclusions about an "uncertainty" statistic. The same is true for "decision making" experiments, but it is less overt. This is easiest to understand with an example.

Imagine you are trying to decide if you want to bring an umbrella with you to work. An umbrella is annoying to bring with you, so you only want to pack it if the chance of rain is greater than 10%. Unfortunately, your weather prediction app only provides you with the predicted daily rainfall. Therefore, your decision will be improved with the inclusion of uncertainty. This is *not* because uncertainty in general is important for decision making, but because it gives you the tools required to calculate the *actual* statistic you are basing your decision on. In this sense, uncertainty is no more "special" to decision making than weight special to a BMI calculation.

This means the uncertainty visualisations that would perform the best in decision making would simply display the uncertainty statistic we are interested in, such as the variance, or probability of an event, using existing visualisation principles. This is precisely what we observe in the literature. @fig-exceed depicts an exceedance probability map that has was designed as an alternative to the choropleth map to improve decision making under uncertainty [@Kuhnert2018; @Lucchesi2021]. A keen viewer may notice that the "exceedance probability map" is actually just a choropleth map, only the statistic being displayed has changed. We are not sure it is productive to categorise this visualisation as an uncertainty visualisation.

@@

There seem to be two different definitions of "uncertainty visualisation" floating around in the literature. The first considers *any* visualisation of error, variance, or probability to be an uncertainty visualisation. The second believes an "uncertainty visualisation" is the output of a function that takes a normal visualisation as an input, and transforms it to include uncertainty information. The former group believe the purpose of uncertainty visualisation to provide signal about a distribution, while the later believe it should act as noise to obfuscate a signal. The lack of explicit distinction between these two motivations leaves the literature muddled and reviewers struggle to understand if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014]. This disagreement creates constant contradictions in what the literature considers to be an "uncertainty visualisation". For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, and @Wickham2011 suggests their product plot framework, which includes histograms and bar charts, should be extended to include uncertainty. Additionally, at least one of these charts have been used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. If you view uncertainty as a function applied to an existing graphic, then you would believe a pie chart and bar chart are not uncertainty visualisations, as they are yet to have the "uncertainty visualisation function" applied to them. If you view uncertainty as any graphic that depicts an "uncertainty statistic" then there are no limitations on which graphics can or cannot be uncertainty visualisations.

@@

It is clear that the pixel-map is not only "suppressing" the false information, but it is doing so by simulating *more* information. The efficacy of this method means that visualisations of simulated samples pop up repeatedly in the literature, with examples including samples that are animated over time [@Hullman2015; @Blenkinsop2000], pixel-maps, and spaghetti time series plots. Not only does this method help readers understand the plot level "gist", it is also unlikely to damage the viewers ability to extract individual estimates. Extracting global statistics, such as the mean or variance, from a sample can be done with relative each, especially when those values are mapped to color [@Franconeri2021]. Additionally, @Ndlovu2023 found that participants applied the same methods they used for simple choropleth maps to complicated uncertainty maps even if that take away was invalid. The pixel map can be read the same way as a choropleth map, which allows us to leverage that automatic response. Therefore, the pixel map performs signal suppression, without sacrificing the viewers ability to extract general statistics, unless those statistics *should* be harder to extract due to the uncertainty in the value. So is this the best uncertainty visualisation? And if so, why?
